---
title: "Central Limit Theorem Simulation Using Exponential Distribution"
author: "Ruel Barredo Laranjo"
date: "March 2026 (Updated from 2022 Coursework)"
output: html_document
---


```r

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(123)

```



## Introduction

This project investigates the Central Limit Theorem (CLT) using simulations of the exponential distribution.

We simulate 1000 samples of 40 exponentially distributed random variables with:

λ (rate parameter) = 0.2

Theoretical mean = 1/λ = 5

Theoretical variance = (1/λ²)/n

The objective is to:

Compare sample mean to theoretical mean

Compare sample variance to theoretical variance

Demonstrate approximate normality of sample means

🔧 ## Simulation Setup

```{r simulation}
lambda <- 0.2
n <- 40
num_sim <- 1000

set.seed(123)

sample_means <- replicate(num_sim, mean(rexp(n, lambda)))

theoretical_mean <- 1 / lambda
theoretical_variance <- (1 / lambda^2) / n
```


📊 2. Variance Comparison
theoretical_variance <- (1 / lambda^2) / n
sample_variance <- var(sample_means)

theoretical_variance
sample_variance
Interpretation

The simulated variance aligns closely with the theoretical variance predicted by the Central Limit Theorem.

📉 3. Demonstrating Normality

```{r}
hist(sample_means,
     probability = TRUE,
     main = "Sampling Distribution of the Mean (n = 40)",
     xlab = "Sample Mean",
     col = "#4C72B0",
     border = "white")

curve(dnorm(x,
            mean = theoretical_mean,
            sd = sqrt(theoretical_variance)),
      col = "red",
      lwd = 2,
      add = TRUE)

legend("topright",
       legend = c("Histogram", "Normal Curve"),
       col = c("#4C72B0", "red"),
       lwd = 2)
```
     
     ## 4. Statistical Test of Normality


```{r}
shapiro.test(sample_means) 

```
      
## Interpretation

The simulated sample mean closely approximates the theoretical mean (5), confirming the Law of Large Numbers.
The theoretical variance of the sampling distribution is also consistent with the observed sample variance, supporting the expected behavior described by the Central Limit Theorem.
Although the Shapiro–Wilk test detects a slight deviation from perfect normality (p < 0.05), this is expected with a large number of simulations, as the test becomes highly sensitive. The histogram, normal curve overlay, and overall bell-shaped symmetry indicate that the distribution of sample means is approximately normal.
This demonstrates that even when the underlying exponential distribution is right-skewed, the sampling distribution of the mean becomes approximately normal as the sample size increases, consistent with the Central Limit Theorem.


🔄 Comparison: Raw Exponential vs Sample Means
raw_exp <- rexp(1000, lambda)

par(mfrow = c(1,2))

hist(raw_exp,
     main = "Raw Exponential Distribution",
     col = "salmon")

hist(sample_means,
     main = "Sample Means Distribution",
     col = "lightgreen")
Interpretation

The exponential distribution is highly skewed.

The distribution of sample means is approximately symmetric and bell-shaped.

This visual comparison strongly supports CLT.


✅  Conclusion

This simulation demonstrates the key principles of the Central Limit Theorem:
The sample mean converges to the theoretical mean as the number of simulations increases.
The sample variance approaches the theoretical variance of the sampling distribution.
Even though the original exponential distribution is highly right-skewed, the sampling distribution of the mean becomes approximately normal as the sample size increases.
These results reinforce foundational statistical theory that is essential in data science, particularly for inference, hypothesis testing, and predictive modeling.

```{r}
